---
title: "Data Analysis Examples with PSInetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Analysis Examples with PSInetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE since we don't have actual data during package build
)
```

```{r setup}
library(PSInetR)
library(DBI)
library(duckdb)
library(dplyr)
library(dbplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
```

## Introduction

This vignette provides practical examples for analyzing plant water potential data using the PSInetR package. These examples demonstrate common analysis workflows and show how to integrate PSInet data with other R packages for ecological research.

Before working through these examples, make sure you have downloaded the PSInet data using `get_psi_data()`. See the "Getting Started with PSInetR" and "Working with DuckDB" vignettes for details on downloading the data, understanding database structure, and accessing the database.

## Example 1: Basic Data Exploration

Let's start by exploring the database structure and getting familiar with the data:

```{r basic-exploration}
# Connect to the DuckDB database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# List available tables
dbListTables(con)

# Preview the study sites
study_sites <- tbl(con, "study_site") |>
  collect()
head(study_sites)

# Count plants by species
species_counts <- tbl(con, "plant") |>
  group_by(genus, specific_epithet) |>
  summarize(plant_count = n()) |>
  arrange(desc(plant_count)) |>
  collect()
head(species_counts)

# Don't forget to disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Example 2: Exploring Water Potential Data

Now let's dive into the water potential measurements for a specific species:

```{r water-potential}
# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Get water potential data for a specific plant - Quercus alba
quercus_ids <- tbl(con, "plant") |> 
  filter(genus == "Quercus") 

# right join by filtered dataset above
wp_quercus <- tbl(con, "chamber_wp") |> 
  right_join(quercus_ids, by = c("dataset_name", "individual_id", "plot_id"))

# collect just to check and see dimensions
wp_quercus|> collect()

# Plot water potential vs leaf area
wp_quercus |> 
  collect() |>
  drop_na(leaf_area_index_m2_m2) |> 
  ggplot(aes(x = leaf_area_index_m2_m2, y = water_potential_mean)) +
  geom_point() +
  labs(title = "Quercus Water Potential by Leaf Area",
       x = "Leaf Area (m2)",
       y = "Water Potential (MPa)") +
  theme_minimal()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Example 3: Combining Environmental and Water Potential Data

This example shows how to join water potential data with meteorological variables:

```{r combined-analysis}
# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Get plants, water potential, and environmental data for a study site
study <- "Ben_1"

# Get water potential data
wp_data <- tbl(con, "chamber_wp") |>
  filter(dataset_name == study) |>
  collect()

# Get meteorological data
met_data <- tbl(con, "met_var") |>
  filter(dataset_name == study) |>
  collect()

# Join water potential and VPD data by date and time
combined_data <- wp_data |>
  inner_join(met_data, by = c("dataset_name", "date", "time")) |>
  select(date, water_potential_mean, vapor_pressure_deficit_k_pa, 
         air_temperature_c, precipitation_mm)

# Plot water potential vs. VPD
combined_data |>
  ggplot(aes(x = vapor_pressure_deficit_k_pa, y = water_potential_mean)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Plant Water Potential vs. Vapor Pressure Deficit",
       x = "VPD (kPa)",
       y = "Water Potential (MPa)") +
  theme_minimal()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Example 4: Time-of-Day Analysis

This example demonstrates how to analyze diurnal patterns in water potential measurements:

```{r time-analysis}
# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Get predawn and midday water potential measurements
wp_data <- tbl(con, "chamber_wp") |>
  filter(
    organ == "stem" | organ == "leaf",
    !is.na(water_potential_mean)
  ) |>
  left_join(
    tbl(con, "plant"),
    by = c("dataset_name", "individual_id")
  ) |>
  collect()

# Create time of day category based on time values
# The time column appears to contain seconds since midnight
# Convert to hours for easier categorization
wp_data_filtered <- wp_data |>
  mutate(
    # Convert character time to numeric seconds, then to hours
    time_numeric = seconds(hms(time)),
    hour_decimal = time_numeric / 3600,  # Convert seconds to hours
    hour = floor(hour_decimal),          # Extract integer hour
    minute = round((hour_decimal - hour) * 60), # Extract minutes
    
    # Categorize based on typical measurement times
    # NOTE: Adjust these ranges based on your specific measurement protocols
    time_of_day = case_when(
      hour >= 4 & hour <= 7 ~ "Predawn",    # 4:00 AM to 7:59 AM
      hour >= 11 & hour <= 14 ~ "Midday",   # 11:00 AM to 2:59 PM
      TRUE ~ "Other"
    )
  ) |>
  filter(time_of_day %in% c("Predawn", "Midday"))

# Alternative approach if you want to see what times are actually present:
# wp_data |>
#   mutate(
#     time_numeric = seconds(hms(time)),
#     hour_decimal = time_numeric / 3600,
#     hour = floor(hour_decimal),
#     minute = round((hour_decimal - hour) * 60),
#     clock_time = sprintf("%02d:%02d", hour, minute)
#   ) |>
#   count(clock_time, sort = TRUE) |>
#   head(20)  # This will show you the most common measurement times

# Visualize water potential patterns by time of day and species
wp_data_filtered |>
  add_count(genus) |>
  filter(n >= 500) |>  # Adjust threshold as needed
  ggplot(aes(x = water_potential_mean, y = genus, fill = time_of_day)) +
  geom_boxplot() +
  labs(
    title = "Water Potential by Species and Time of Day",
    x = "Water Potential (MPa)",
    y = "Genus",
    fill = "Time of Day"
  ) +
  theme_minimal()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Example 5: Multi-table Analysis - Combining Plant and Site Information

This example demonstrates how to combine data from multiple tables to analyze water potential measurements with species and study site context:

```{r multi-table}
# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Simple multi-table join: water potential + plant species + study sites
multi_table_data <- tbl(con, "chamber_wp") |>
  # Join with plant table to get species information
  inner_join(tbl(con, "plant"), by = c("dataset_name", "individual_id")) |>
  # Join with study site table to get location information  
  inner_join(tbl(con, "study_site"), by = "dataset_name") |>
  # Filter for good quality data
  filter(!is.na(water_potential_mean), !is.na(genus)) |>
  # Select key columns
  select(dataset_name, genus, specific_epithet, water_potential_mean, 
         organ, latitude_wgs84, begin_year, end_year) |>
  collect()

# Summarize by genus across different studies
genus_summary <- multi_table_data |>
  group_by(genus) |>
  summarize(
    mean_wp = mean(water_potential_mean, na.rm = TRUE),
    n_measurements = n(),
    n_studies = n_distinct(dataset_name),
    .groups = "drop"
  ) |>
  # Focus on genera with data from multiple studies
  filter(n_studies >= 2, n_measurements >= 20) |>
  arrange(mean_wp)

# Plot 1: Water potential by genus (for genera in multiple studies)
genus_summary |>
  ggplot(aes(x = reorder(genus, mean_wp), y = mean_wp)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = paste0("n=", n_measurements)), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(
    title = "Mean Water Potential by Genus",
    subtitle = "Genera with data from multiple studies (≥20 measurements)",
    x = "Genus",
    y = "Mean Water Potential (MPa)"
  ) +
  theme_minimal()

# Plot 2: Study timeline showing data collection periods
study_timeline <- multi_table_data |>
  select(dataset_name, begin_year, end_year, latitude_wgs84) |>
  distinct() |>
  filter(!is.na(begin_year), !is.na(end_year)) |>
  mutate(
    study_duration = end_year - begin_year + 1,
    latitude_group = ifelse(latitude_wgs84 >= 35, "Northern", "Southern")
  )

study_timeline |>
  # filter for reasonable subset
  filter(study_duration >= 6 | begin_year >= 2020) |>
  ggplot(aes(x = begin_year, xend = end_year, 
             y = reorder(dataset_name, begin_year), yend = reorder(dataset_name, begin_year),
             color = latitude_group)) +
  geom_segment(linewidth = 2, alpha = 0.7) +
  geom_point(aes(x = begin_year), size = 0.5) +
  geom_point(aes(x = end_year), size = 0.5) +
  labs(
    title = "Study Timeline and Geographic Distribution",
    x = "Year",
    y = "Study",
    color = "Region"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# Print summary
cat("Multi-table join results:\n")
cat("Total water potential measurements:", nrow(multi_table_data), "\n") 
cat("Number of genera:", length(unique(multi_table_data$genus)), "\n")
cat("Number of studies:", length(unique(multi_table_data$dataset_name)), "\n")
cat("Year range:", min(multi_table_data$begin_year, na.rm = TRUE), "-", 
    max(multi_table_data$end_year, na.rm = TRUE), "\n")

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Example 6: Seasonal Patterns

Analyze how water potential varies throughout the growing season:

```{r seasonal}
# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Extract seasonal patterns
seasonal_data <- tbl(con, "chamber_wp") |>
  inner_join(tbl(con, "plant"), by = c("dataset_name", "individual_id")) |>
  filter(!is.na(water_potential_mean)) |>
  collect() |>
  mutate(
    # Extract date components
    year = year(date),
    month = month(date),
    day = day(date),
    # Create day of year
    date_parsed = as.Date(paste(year, month, day, sep = "-")),
    day_of_year = as.numeric(format(date_parsed, "%j"))
  ) |>
  filter(!is.na(day_of_year))

# Plot seasonal patterns - this plots a lot of data and takes a while
seasonal_data |>
  ggplot(aes(x = day_of_year, y = water_potential_mean)) +
  geom_hex() + 
  # geom_smooth(method = "loess", se = TRUE) +
  # facet_wrap(~genus, scales = "free_y") + # could also facet by genus
  labs(
    title = "Seasonal Patterns in Water Potential",
    x = "Day of Year",
    y = "Water Potential (MPa)"
  ) +
  theme_minimal()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Advanced Integration Examples

### Working with Spatial Data

```{r spatial, eval=FALSE}
# If you have spatial analysis needs
library(sf)

# Connect to database
con <- dbConnect(duckdb::duckdb(), get_db_path())

# Get study site locations
site_locations <- tbl(con, "study_site") |>
  filter(!is.na(latitude_wgs84), !is.na(longitude_wgs84)) |>
  collect() |>
  st_as_sf(coords = c("longitude_wgs84", "latitude_wgs84"), crs = 4326)

# Now you can use spatial analysis functions
dbDisconnect(con, shutdown = TRUE)
```

### Time Series Analysis

```{r timeseries, eval=FALSE}
# For time series analysis
library(lubridate)
library(forecast)

con <- dbConnect(duckdb::duckdb(), get_db_path())

# Get time series data for a specific site/species
ts_data <- tbl(con, "chamber_wp") |>
  inner_join(tbl(con, "plant"), by = c("dataset_name", "individual_id")) |>
  filter(dataset_name == "Ada_1", genus == "Thuja") |>
  collect() |>
  mutate(
    date_time = ymd(date) + seconds(hms(time))
  ) |>
  arrange(date_time)

# Create a complete time series by aggregating to daily means
daily_wp <- ts_data |>
  mutate(date_only = as.Date(date_time)) |>
  group_by(date_only) |>
  summarize(
    mean_wp = mean(water_potential_mean, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  ) |>
  arrange(date_only)

# Plot the time series
ggplot(daily_wp, aes(x = date_only, y = mean_wp)) +
  geom_line() +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(
    title = "Daily Mean Water Potential Time Series",
    x = "Date",
    y = "Water Potential (MPa)"
  ) +
  theme_minimal()

# Autocorrelation analysis
acf_result <- acf(daily_wp$mean_wp, na.action = na.pass, plot = TRUE,
                  main = "Autocorrelation of Water Potential")

# Decompose the time series if there's enough data
if (nrow(daily_wp) >= 365) {
  # Create regular time series object
  ts_wp <- ts(daily_wp$mean_wp, frequency = 365)
  
  # Decompose into trend, seasonal, and random components
  decomp <- decompose(ts_wp, type = "additive")
  
  # Plot decomposition
  plot(decomp, main = "Time Series Decomposition")
}

dbDisconnect(con, shutdown = TRUE)
```

## Best Practices for Analysis

1. **Always disconnect from the database** when finished with your analysis
2. **Use `collect()` strategically** - filter data in the database before collecting to R
3. **Handle missing values** appropriately for your analysis
4. **Check data quality** using the flag tables when available
5. **Document your analysis** including any data filtering decisions
6. **Consider temporal and spatial scales** when interpreting results

## Next Steps

For more advanced database operations and performance optimization, see the "Working with DuckDB" vignette. The PSInet database offers rich opportunities for:

- Cross-species comparisons of drought responses
- Climate change impact assessments
- Hydraulic model parameterization
- Ecosystem-scale water cycling studies

Remember to cite the appropriate data contributors when using PSInet data in your research. Dataset attribution information can be found in the `study_site` table.
