---
title: "Getting Started with PSInetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PSInetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PSInetR)
```

## Introduction

PSInetR provides functions to access and work with plant water potential (Ψ) data from the PSInet Research Coordination Network. This vignette demonstrates how to download, access, and analyze this data.

PSInet is a network focused on collecting and aggregating plant water potential measurements to improve understanding of ecosystem responses to drought and heat stress. Water potential can be thought of as the "blood pressure of the natural world" that directly controls plant functioning during drought. The PSInetR package gives you access to this valuable dataset.

## Downloading the Data

The plant water potential data can be downloaded from either a repository or Zenodo (forthcoming). By default, the data is downloaded as a DuckDB file, but it can also be downloaded as CSV files.

```{r downloading, eval=FALSE}
# Download data from repository as DuckDB (default)
get_psi_data()

# Download individual CSV files from Zenodo (forthcoming)
# get_psi_data(source = "zenodo", format = "csv")

# If the repository is private, provide a GitHub token
get_psi_data(github_token = "your_github_token")

# Alternatively, store your GitHub token in .Renviron file:
# GITHUB_PAT=your_github_token_here
# Then simply call:
# get_psi_data()
```


## PSInet Database Structure

The PSInet database follows a hierarchical structure that organizes water potential data and relevant contextual information. For detailed descriptions of all variables and their meanings, see the [PSInetDB data schema documentation](https://github.com/PSInetRCN/PSInetDB/blob/main/docs/schema/data_schema.md).

**Important Note**: Many variables like `plot_id` and `individual_id` are author-provided and were not harmonized across studies. These identifiers will vary significantly in format and naming conventions between different research groups and datasets.

### Core Tables

1. **study_site**: Contains information about research locations where plant-soil interaction studies are conducted, including geographic coordinates, study period, and investigator details.

2. **data_description**: Provides descriptions of the available data types and measurements for each study site.

3. **addtl_data**: Information about additional data availability across networks.

4. **treatment**: Documents experimental treatments applied at study sites.

5. **plot**: Contains information about experimental plots, including vegetation type, soil properties, and terrain details.

6. **plant**: Stores data about individual plants or groups of plants, including species information, size measurements, and plant social status.


### Measurement Tables

1. **chamber_wp**: Manual pressure chamber water potential measurements.

2. **auto_wp_sensor**: Sensor details associatd with automated measurements.

3. **auto_wp**: Automated water potential sensor measurements.

4. **soil_var**: Soil water content and soil water potential measurements.

5. **met_var**: Meteorological variables like precipitation, air temperature, relative humidity, and radiation.

### Auxillary Tables

For cross-network compatibility with SAPFLUXNET:
- **sapfluxnet**

Each measurement table has a corresponding flag table for quality control:
- **chamber_wp_flag**, **auto_wp_sensor_flag**, **auto_wp_flag**, **soil_var_flag**, **met_var_flag**

The plant table also has a corresponding flag table:
- **plant_flag**

### Key Database Relationships

```
study_site → plot → plant → water potential measurements
     |        |       |
     |        |       → chamber_wp (manual measurements)
     |        |       → auto_wp (automated measurements)
     |        |
     |        → soil_var (soil measurements)*
     |
     → met_var (meteorological variables)
     → treatment (experimental treatments)
     → addtl_data (additional data links)
```

*Note: Soil measurements (`soil_var`) can be associated with either plots (including whole study sites) or individual plants, depending on the measurement approach used by each research group.

## Working with the Data

Once you have downloaded the data, you can begin analyzing it. The PSInet database is stored in DuckDB format, which provides fast and efficient querying capabilities.

### Basic Database Connection

```{r basic-connection, eval=FALSE}
library(DBI)
library(duckdb)

# Get the path to the database and connect
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# List all available tables
tables <- dbListTables(con)
print(tables)

# Always disconnect when finished
dbDisconnect(con, shutdown = TRUE)
```

### Next Steps

For detailed analysis examples and workflows, see the "Data Analysis Examples with PSInetR" vignette:

```{r next-steps, eval=FALSE}
# View the examples vignette
vignette("data-analysis-examples", package = "PSInetR")
```

For advanced database operations and performance optimization, refer to the "Working with DuckDB" vignette:

```{r advanced, eval=FALSE}
# View the DuckDB vignette
vignette("working-with-duckdb", package = "PSInetR")
```

## Research Applications

As the PSInet database grows, researchers can use it for:

1. **Drought Response Studies**: Compare how different species respond to water stress
2. **Climate Change Research**: Analyze water potential patterns under changing climate conditions
3. **Hydraulic Modeling**: Use water potential data to parameterize plant hydraulic models
4. **Ecosystem Monitoring**: Track ecosystem water status across different environments
5. **Cross-Network Synthesis**: Integrate with other ecological networks through linked data

## Conclusion

This vignette provided an introduction to PSInetR and how to download and access the PSInet database. The package makes plant water potential data easily accessible for ecological research, drought studies, and plant physiology investigations.

The PSInet database represents a valuable resource for understanding plant-water relationships across diverse ecosystems and species. By providing standardized access to this data, PSInetR enables researchers to conduct large-scale synthetic analyses that would not be possible with individual datasets alone.
