---
title: "Getting Started with PSINetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PSINetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PSINetR)
```

## Introduction

PSINetR provides functions to access and work with plant water potential data. This vignette demonstrates how to download, access, and analyze this data.

## Downloading the Data

The plant water potential data can be downloaded from either a repository or Zenodo. By default, the data is downloaded as a DuckDB file, but it can also be downloaded as CSV files.

```{r downloading, eval=FALSE}
# Download data from repository as DuckDB (default)
get_psi_data()

# Download data from Zenodo as CSV files
get_psi_data(source = "zenodo", format = "csv")

# If the repository is private, provide a GitHub token
get_psi_data(github_token = "your_github_token")
```

## Checking if Data is Available

Before attempting to use the data, you can check if it's available locally.

```{r checking, eval=FALSE}
# Check if DuckDB data is available
check_psi_data()

# Check if CSV data is available
check_psi_data(format = "csv")

# Check and validate the data
check_psi_data(validate = TRUE)
```

## Working with the Data

After downloading the data, you can start working with it. The following examples demonstrate basic data analysis workflows.

### Example 1: Basic Data Exploration

[Example code for data exploration will be added here]

### Example 2: Visualizing Plant Water Potential Data

[Example code for data visualization will be added here]

### Example 3: Advanced Analysis

[Example code for advanced analysis will be added here]

## Conclusion

This vignette provided a basic introduction to the PSINetR package. For more detailed information on specific topics, refer to the other vignettes and the function documentation.