---
title: "Getting Started with PSInetR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PSInetR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PSInetR)
```

## Introduction

PSInetR provides functions to access and work with plant water potential (Ψ) data from the PSInet Research Coordination Network. This vignette demonstrates how to download, access, and analyze this data.

PSInet is a network focused on collecting and aggregating plant water potential measurements to improve understanding of ecosystem responses to drought and heat stress. Water potential can be thought of as the "blood pressure of the natural world" that directly controls plant functioning during drought. The PSInetR package gives you access to this valuable dataset.

## Downloading the Data

The plant water potential data can be downloaded from either a repository or Zenodo. By default, the data is downloaded as a DuckDB file, but it can also be downloaded as CSV files.

```{r downloading, eval=FALSE}
# Download data from repository as DuckDB (default)
get_psi_data()

# Download individual CSV files from Zenodo
get_psi_data(source = "zenodo", format = "csv")

# If the repository is private, provide a GitHub token
get_psi_data(github_token = "your_github_token")
```

## Checking if Data is Available

Before attempting to use the data, you can check if it's available locally.

```{r checking, eval=FALSE}
# Check if DuckDB data is available
check_psi_data()

# Check if CSV data is available
check_psi_data(format = "csv")

# Check and validate the data
check_psi_data(validate = TRUE, dir = "/Users/keatonwilson/Documents/Projects/PSInetR/vignettes/")
```

## PSInet Database Structure

The PSInet database follows a hierarchical structure that organizes water potential data and relevant contextual information:

### Core Tables

1. **study_site**: Contains information about research locations where plant-soil interaction studies are conducted, including geographic coordinates, study period, and investigator details.

2. **data_description**: Provides descriptions of the available data types and measurements for each study site.

3. **treatments**: Documents experimental treatments applied at study sites.

4. **plots**: Contains information about experimental plots, including vegetation type, soil properties, and terrain details.

5. **plants**: Stores data about individual plants or groups of plants, including species information, size measurements, and plant social status.

### Measurement Tables

1. **press_chamb_wp**: Manual pressure chamber water potential measurements.

2. **auto_wp**: Automated water potential sensor measurements.

3. **soil_moisture**: Soil water content and soil water potential measurements.

4. **env_vars**: Environmental variables like precipitation, temperature, humidity, and radiation.

### Key Database Relationships

```
study_site → plots → plants → water potential measurements
     |         |        |
     |         |        → automated water potential measurements
     |         |
     |         → soil moisture measurements
     |
     → environmental variables
```

## Working with the Data

After downloading the data, you can start working with it. The following examples demonstrate basic data analysis workflows.

### Example 1: Basic Data Exploration

```{r basic-exploration, eval=FALSE}
library(DBI)
library(duckdb)
library(dplyr)
library(dbplyr)

# Connect to the DuckDB database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# List available tables
dbListTables(con)

# Preview the study sites
study_sites <- tbl(con, "study_site") |>
  collect()
head(study_sites)

# Count plants by species
species_counts <- tbl(con, "plants") |>
  group_by(genus, specific_epithet) |>
  summarize(plant_count = n()) |>
  arrange(desc(plant_count)) |>
  collect()
head(species_counts)

# Don't forget to disconnect
dbDisconnect(con, shutdown = TRUE)
```

### Example 2: Exploring Water Potential Data

```{r water-potential, eval=FALSE}
library(DBI)
library(duckdb)
library(dplyr)
library(dbplyr)
library(ggplot2)

# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Get water potential data for a specific plant - Sorghum bicolor
sorghum_ids <- tbl(con, "plant_types") |> 
  filter(genus == "Sorghum" & specific_epithet == "bicolor") 

# right join by filtered dataset above
wp_sorg <- tbl(con, "press_chamb_wp") |> 
  right_join(sorghum_ids, by = c("dataset_name", "individual_id"))

# collect just to check and see dimensions
wp_sorg |> collect()

# Plot water potential vs leaf area
wp_sorg |> 
  ggplot(aes(x = leaf_area_index_m2_m2, y = water_potential_mean)) +
  geom_point() +
  labs(title = "Sorghum Water Potential by Leaf Area",
       x = "Leaf Area (m2)",
       y = "Water Potential (MPa)") +
  theme_minimal()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

### Example 3: Combining Environmental and Water Potential Data

```{r combined-analysis, eval=FALSE}
library(DBI)
library(duckdb)
library(dplyr)
library(ggplot2)

# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Get plants, water potential, and environmental data for a study site
study <- "Ben_1"

# Get water potential data
wp_data <- tbl(con, "press_chamb_wp") |>
  filter(dataset_name == study) |>
  collect()

# Get environmental data
env_data <- tbl(con, "env_vars") |>
  filter(dataset_name == study) |>
  collect()

# Join water potential and VPD data by date and time
combined_data <- wp_data |>
  inner_join(env_data, by = c("dataset_name", "date", "time")) |>
  select(date, water_potential_mean, vapor_pressure_deficit_k_pa, 
         air_temperature_c, precipitation_mm)

# Plot water potential vs. VPD
combined_data |>
  ggplot(aes(x = vapor_pressure_deficit_k_pa, y = water_potential_mean)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Plant Water Potential vs. Vapor Pressure Deficit",
       x = "VPD (kPa)",
       y = "Water Potential (MPa)") +
  theme_minimal()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Adding the Database to Your Analysis Pipeline

The PSInet database can be integrated with other R packages for advanced analysis:

```{r integration, eval=FALSE}
# Example: Use tidyverse for data wrangling and visualization
library(dplyr)
library(ggplot2)
library(DBI)
library(duckdb)

# Connect to the database
db_path <- get_db_path()
con <- dbConnect(duckdb::duckdb(), db_path)

# Get predawn and midday water potential measurements
wp_data <- tbl(con, "press_chamb_wp") |>
  filter(
    organ == "stem" | organ == "leaf",
    !is.na(water_potential_mean)
  ) |>
  left_join(
    tbl(con, "plants"),
    by = c("dataset_name", "individual_id")
  ) |>
  collect()

# Create time of day category based on time values
# The time column appears to contain seconds since midnight
# Convert to hours for easier categorization
wp_data_filtered <- wp_data |>
  mutate(
    # Convert character time to numeric seconds, then to hours
    time_numeric = as.numeric(time),
    hour_decimal = time_numeric / 3600,  # Convert seconds to hours
    hour = floor(hour_decimal),          # Extract integer hour
    minute = round((hour_decimal - hour) * 60), # Extract minutes
    
    # Categorize based on typical measurement times
    # NOTE: Adjust these ranges based on your specific measurement protocols
    time_of_day = case_when(
      hour >= 4 & hour <= 7 ~ "Predawn",    # 4:00 AM to 7:59 AM
      hour >= 11 & hour <= 14 ~ "Midday",   # 11:00 AM to 2:59 PM
      TRUE ~ "Other"
    )
  ) |>
  filter(time_of_day %in% c("Predawn", "Midday"))

# Alternative approach if you want to see what times are actually present:
# wp_data |>
#   mutate(
#     time_numeric = as.numeric(time),
#     hour_decimal = time_numeric / 3600,
#     hour = floor(hour_decimal),
#     minute = round((hour_decimal - hour) * 60),
#     clock_time = sprintf("%02d:%02d", hour, minute)
#   ) |>
#   count(clock_time, sort = TRUE) |>
#   head(20)  # This will show you the most common measurement times

# Visualize water potential patterns by time of day and species
wp_data_filtered %>%
  ggplot(aes(x = genus, y = water_potential_mean, fill = time_of_day)) +
  geom_boxplot() +
  labs(
    title = "Water Potential by Species and Time of Day",
    x = "Genus",
    y = "Water Potential (MPa)",
    fill = "Time of Day"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Future Directions

As the PSInet database grows, researchers can:

1. **Compare Drought Responses**: Evaluate how different species respond to water stress
2. **Assess Climate Change Impacts**: Analyze how water potential patterns change with climate variables
3. **Build Hydraulic Models**: Use water potential data as inputs for plant hydraulic models
4. **Improve Drought Forecasting**: Develop better predictive models of ecosystem responses to drought

## Conclusion

This vignette provided a basic introduction to the PSInetR package and the PSInet database structure. The package makes it easy to access and analyze plant water potential data for ecological research, drought studies, and plant physiology investigations.

For more detailed information on working specifically with the DuckDB database format, refer to the "working-with-duckdb" vignette.
