---
title: "Working with Plant Water Potential Data in DuckDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Plant Water Potential Data in DuckDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE since we don't have actual data during package build
)
```

```{r setup}
library(PSINetR)
library(dplyr)
library(DBI)
library(duckdb)
```

## Introduction

This vignette demonstrates how to work with plant water potential data stored in DuckDB format. DuckDB is a high-performance analytical database system that works well with R and provides efficient querying capabilities for plant water potential (Ψ) datasets.

### Why DuckDB?

DuckDB offers several advantages for working with plant water potential data:

1. **Performance**: DuckDB provides fast analytical queries, even on larger datasets
2. **Ease of use**: No separate server installation required
3. **SQL and dplyr compatibility**: Familiar query syntax options
4. **Zero configuration**: Works immediately with R
5. **Column-oriented storage**: Optimized for analytical workloads common in ecological research

## Connecting to the Database

First, let's download the DuckDB data (if you haven't already) and connect to it:

```{r connect}
# Download the data (if not already available)
data_path <- get_psi_data(format = "duckdb")

# Connect to the DuckDB database
con <- dbConnect(duckdb::duckdb(), data_path)

# List available tables
tables <- dbListTables(con)
print(tables)
```

## Understanding the Database Structure

The PSInet database contains several tables organized in a hierarchical structure. Let's first explore what tables are available:

```{r explore_tables}
# List all tables in the database
dbListTables(con)

# Look at the database schema
db_schema <- DBI::dbGetQuery(con, "
  SELECT 
    table_name, 
    column_name, 
    data_type
  FROM information_schema.columns 
  WHERE table_schema = 'main'
  ORDER BY table_name, ordinal_position
")

# View the schema of a specific table
press_chamb_fields <- dbListFields(con, "press_chamb_wp")
print(press_chamb_fields)
```

## Basic Querying

Now that we understand the database structure, let's explore the data with some basic queries:

```{r basic_queries}
# Get a list of all study sites
study_sites <- dbGetQuery(con, "SELECT dataset_name, latitude_wgs84, longitude_wgs84 FROM study_site LIMIT 10")
print(study_sites)

# Get plant species information
plant_species <- dbGetQuery(con, "SELECT DISTINCT genus, specific_epithet FROM plants LIMIT 10")
print(plant_species)

# Get statistics on water potential measurements
wp_stats <- dbGetQuery(con, "
  SELECT 
    COUNT(*) as count,
    AVG(water_potential_mean) as avg_potential,
    MIN(water_potential_mean) as min_potential,
    MAX(water_potential_mean) as max_potential
  FROM press_chamb_wp
  WHERE water_potential_mean IS NOT NULL
")
print(wp_stats)
```

## Using dplyr with DuckDB

You can also use dplyr syntax to work with the database, which often makes complex queries more readable:

```{r dplyr_queries}
# Create references to the tables
press_chamb_wp_tbl <- tbl(con, "press_chamb_wp")
plants_tbl <- tbl(con, "plants")
plots_tbl <- tbl(con, "plots")

# Query using dplyr syntax for water potential by species
wp_by_species <- press_chamb_wp_tbl %>%
  inner_join(plants_tbl, by = c("dataset_name", "individual_id")) %>%
  group_by(genus, specific_epithet) %>%
  summarize(
    count = n(),
    avg_potential = mean(water_potential_mean, na.rm = TRUE),
    min_potential = min(water_potential_mean, na.rm = TRUE),
    max_potential = max(water_potential_mean, na.rm = TRUE)
  ) %>%
  arrange(desc(count)) %>%
  collect()  # This executes the query and brings results into R

print(wp_by_species)

# Find measurements by organ type
organ_summary <- press_chamb_wp_tbl %>%
  group_by(organ, canopy_position) %>%
  summarize(
    count = n(),
    avg_potential = mean(water_potential_mean, na.rm = TRUE),
    avg_n_samples = mean(water_potential_n, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(desc(count)) %>%
  collect()

print(organ_summary)
```

## Advanced Analyses

Here are some examples of more advanced analyses that combine multiple data tables:

```{r advanced_analyses}
# Analyze seasonal patterns in water potential
seasonal_patterns <- press_chamb_wp_tbl %>%
  # Extract month from date (YYYYMMDD format)
  mutate(month = substr(date, 5, 6)) %>%
  group_by(month) %>%
  summarize(
    avg_potential = mean(water_potential_mean, na.rm = TRUE),
    sd_potential = sd(water_potential_mean, na.rm = TRUE),
    sample_count = n()
  ) %>%
  collect()

print(seasonal_patterns)

# Comparing water potential with soil moisture
soil_wp_comparison <- press_chamb_wp_tbl %>%
  # Join with soil moisture data on dataset_name, plot_id, and date
  inner_join(
    tbl(con, "soil_moisture"),
    by = c("dataset_name", "plot_id", "date")
  ) %>%
  # Group by soil moisture categories (using a window function)
  mutate(
    swc_shallow_category = case_when(
      swc_mean_shallow < 0.1 ~ "very low",
      swc_mean_shallow < 0.2 ~ "low",
      swc_mean_shallow < 0.3 ~ "medium",
      TRUE ~ "high"
    )
  ) %>%
  group_by(swc_shallow_category) %>%
  summarize(
    avg_water_potential = mean(water_potential_mean, na.rm = TRUE),
    sd_water_potential = sd(water_potential_mean, na.rm = TRUE),
    n_observations = n()
  ) %>%
  collect()

print(soil_wp_comparison)

# Analyzing the relationship between environmental variables and water potential
env_wp_comparison <- press_chamb_wp_tbl %>%
  # Join with environmental data on dataset_name and date
  inner_join(
    tbl(con, "env_vars"),
    by = c("dataset_name", "date")
  ) %>%
  # Select relevant columns
  select(
    dataset_name, date, water_potential_mean, water_potential_sd,
    vapor_pressure_deficit_k_pa, air_temperature_c, precipitation_mm
  ) %>%
  # Calculate correlations
  summarize(
    vpd_correlation = cor(water_potential_mean, vapor_pressure_deficit_k_pa, 
                           use = "complete.obs"),
    temp_correlation = cor(water_potential_mean, air_temperature_c, 
                            use = "complete.obs"),
    precip_correlation = cor(water_potential_mean, precipitation_mm, 
                              use = "complete.obs"),
    n_observations = n()
  ) %>%
  collect()

print(env_wp_comparison)
```

## Visualization Examples

Let's create some visualizations with the data we've retrieved:

```{r visualizations, message=FALSE, warning=FALSE}
library(ggplot2)
library(lubridate)

# Connect to database (new connection for this section)
con <- dbConnect(duckdb::duckdb(), get_data_dir())

# Get water potential data for visualization
wp_time_data <- tbl(con, "press_chamb_wp") %>%
  filter(!is.na(water_potential_mean)) %>%
  collect() %>%
  mutate(date_parsed = ymd(as.character(date)))  # Convert YYYYMMDD to Date

# Plot water potential over time
ggplot(wp_time_data, aes(x = date_parsed, y = water_potential_mean)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Plant Water Potential Over Time",
    x = "Date",
    y = "Water Potential (MPa)"
  ) +
  theme_minimal()

# Get water potential by species
wp_species <- tbl(con, "press_chamb_wp") %>%
  inner_join(tbl(con, "plants"), by = c("dataset_name", "individual_id", "plot_id")) %>%
  group_by(genus, specific_epithet) %>%
  summarize(
    mean_wp = mean(water_potential_mean, na.rm = TRUE),
    sd_wp = sd(water_potential_mean, na.rm = TRUE),
    count = n()
  ) %>%
  filter(count > 5) %>%  # Only species with enough measurements
  collect()

# Create species name
wp_species$species <- paste(wp_species$genus, wp_species$specific_epithet)

# Plot water potential by species
ggplot(wp_species, aes(x = reorder(species, mean_wp), y = mean_wp)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_wp - sd_wp, ymax = mean_wp + sd_wp), width = 0.2) +
  labs(
    title = "Average Water Potential by Species",
    x = "Species",
    y = "Water Potential (MPa)"
  ) +
  coord_flip() +  # Flip coordinates for better readability
  theme_minimal()

# Don't forget to disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Working with Both Automated and Manual Measurements

The PSInet database contains both manual pressure chamber measurements and automated continuous measurements. Here's how to work with both:

```{r combined_measurements}
# Connect to database
con <- dbConnect(duckdb::duckdb(), get_data_dir())

# Query to combine manual and automated measurements
combined_query <- "
SELECT
  'manual' as measurement_type,
  pc.dataset_name,
  pc.individual_id,
  pc.date,
  pc.time,
  pc.water_potential_mean,
  pc.water_potential_sd,
  p.genus,
  p.specific_epithet
FROM press_chamb_wp pc
JOIN plants p ON pc.dataset_name = p.dataset_name AND pc.individual_id = p.individual_id
WHERE pc.water_potential_mean IS NOT NULL

UNION ALL

SELECT
  'automated' as measurement_type,
  a.dataset_name,
  a.individual_id,
  a.date,
  a.time,
  a.water_potential_mean,
  a.water_potential_sd,
  p.genus,
  p.specific_epithet
FROM auto_wp a
JOIN plants p ON a.dataset_name = p.dataset_name AND a.individual_id = p.individual_id
WHERE a.water_potential_mean IS NOT NULL
LIMIT 1000  -- Limit to avoid memory issues
"

combined_data <- dbGetQuery(con, combined_query)

# Compare manual vs automated measurements
measurement_summary <- combined_data %>%
  group_by(measurement_type) %>%
  summarize(
    count = n(),
    avg_potential = mean(water_potential_mean, na.rm = TRUE),
    sd_potential = sd(water_potential_mean, na.rm = TRUE),
    min_potential = min(water_potential_mean, na.rm = TRUE),
    max_potential = max(water_potential_mean, na.rm = TRUE)
  )

print(measurement_summary)

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Performance Tips

When working with larger PSInet datasets:

1. **Limit data retrieval**: Use `filter()` before `collect()` to reduce data transferred
2. **Use lazy evaluation**: Keep operations in the database as long as possible
3. **Index columns**: If you modify the database, consider adding indices for frequently filtered columns
4. **Close connections**: Always disconnect when done to free resources
5. **Avoid large JOINs in R**: Perform joins in DuckDB with `inner_join()` before collecting

```{r performance_example}
# Example of good practice
con <- dbConnect(duckdb::duckdb(), get_data_dir())

# Efficient query - filtering happens in database
efficient_query <- tbl(con, "press_chamb_wp") %>%
  filter(water_potential_mean < -1.0) %>%
  select(dataset_name, individual_id, date, water_potential_mean) %>%
  collect()

# Disconnect
dbDisconnect(con, shutdown = TRUE)
```

## Conclusion

This vignette demonstrated how to work with the PSInet database in DuckDB format. You've learned how to:

1. Connect to the DuckDB database
2. Explore the database structure
3. Perform basic and advanced queries using SQL and dplyr
4. Join related tables to analyze relationships
5. Visualize water potential data
6. Compare automated and manual measurements

DuckDB provides efficient querying capabilities that work well with both SQL and dplyr syntax, making it a powerful tool for analyzing plant water potential data. As the PSInet database grows, these techniques will help you leverage this valuable resource for ecological research and drought studies.